---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width=8,
  fig.height=8
)
```

```{r}
library(ggplot2)
library(dplyr)
library(wolfeR)
```


# wolfeR

<!-- badges: start -->
<!-- badges: end -->

The goal of wolfeR is to streamline  color schemes for plots of the microbes used in the Wolfe lab at Tufts.

## Installation

You can install the development version of wolfeR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cbedwards/wolfeR")
```

## Useage

We often need to create plots of microbial abundance either aggregated into bar plots or showing time series from experiments. The goal here is streamline the process of using consistent colors for each taxa/species, so that plots have consistent color interpretation across presentations, publications, and working documents. This package provides ready-made access to formalized color palettes in the form of a few functions that integrate with ggplot. 

### generating data

Let's start with some simulated data that resembles real time series data from the Wolfe lab. Here we're imagining we carried out an experiment where we measured the abundance of three species across four time points, using three replicates each. 

```{r}
dat = as.data.frame(expand.grid(name = c("candida", "vibrio",
                                         "brachybacterium"), #species name
                                day = c(1, 3, 7, 14, 21), #day of time series
                                rep = 1:3,#replicate number
                                stringsAsFactors = FALSE))
dat$count = 500 + #baseline
  100000 * (dat$day>1) +  #population generally gets near carrying capacity
  #  after first measurement
  (dat$day>1) * -20000*(dat$name == "candida") +#make candida CC lower
  (dat$day>1) *  30000*(dat$name == "vibrio") + #make vibrio CC higher
  (dat$day>1) * runif(nrow(dat), min = -20000, max = 20000)
# add some random noise to observations except day 1
```

### without wolfeR

ggplot figures default to a single color scheme, which (a) doesn't match Wolfe lab publications, but also (b) is based on the data given. So if we make a plot for one experiment in which we had four species, and another plot for a second experiment with four species, two of which overlapped with the first, there is no guarantee that any of the colors match up between our two plots (even though we would want the two overlapping species to use the same colors).

Here's what the time series looks like for our simulated data: 

```{r}
ggplot(data = dat, aes(x = day,
                       y = count,
                       group = interaction(name, rep),
                       color = name))+
  geom_path(size = 0.8)+
  geom_point(size = 2)+
  scale_y_log10()+
  ggtitle("Simulated time series, 3 replicates per species")

```

And here's what the barplots of that same data would look like, averaging across replicates and separated across days with `facet_wrap()`:

```{r}
dat.summary = dat %>%
  group_by(name, day) %>%
  summarize(count = mean(count))
ggplot(data = dat.summary,
       aes(x = name, y = count, fill = name))+
  geom_col()+
  facet_wrap(~ day)+
  xlab("")+
  ggtitle("Simulated summary data, separated by day of experiment")
```



### Using wolfeR for colors

First, we need to choose which color palette we want. See `help("make_colorvec")` for details; as of the writing of this README, only the color scheme based on Wolfe 2014 (`palette.name = "wolfe2014"`) is implemented. We can visualize a palette with `palette_vis()`.

```{r, fig.height = 7, fig.width = 12}
## showing the default palette
palette_vis("wolfe2014")
```

With the palette in mind, we should check that our data doesn't have any taxa / names that aren't present in the color palette. `palette_check()` is our friend there.

```{r}
# check the palette
palette_check(dat$name, "wolfe2014")
```

As an experiment, let's see what happens if we give it a vector with some non-represented taxa. I'll give a vector of names that are reasonable, plus the species name for common milkweed. 

```{r}
# check the palette
palette_check(c("candida", "vibrio", "brachybacterium", "Asclepias syriaca"), "wolfe2014")
```

KEY POINT: We need *exact* matches between the names in your data and the names in the palette, including capitalization etc. If it seems like the palette should be working, but it's not, check for typos, capitalization, spaces, etc. 

Okay, but when we checked our actual data, with `palette_check(dat$name, "wolfe2014")`, it gave us the all-clear. 

Now we can re-create our figures using the new color palette. If we're working with points and lines, `scale_color_wolfe()` will (generally) add the appropriate colors (for some non-default types of points, we may need to use fill instead). If we're working with bars etc, `scale_fill_wolfe()` will add the appropriate colors. 

These functions both need (a) the vector of taxa names from the data being used (the `dat.names` argument), and (b) the name of the color palette (the `palette.name` argument, defaults to `"wolfe2014"`. (a) seems silly, but is actually really important, because if ggplot gets a list of colors that includes taxa not in our data, it behaves poorly (it assumes we want a color gradient instead, and gives gray colors).

```{r}
ggplot(data = dat, aes(x = day,
                       y = count,
                       group = interaction(name, rep),
                       color = name))+
  geom_path(size = 0.8)+
  geom_point(size = 2)+
  scale_y_log10()+
  scale_color_wolfe(dat.names = dat$name, palette.name = "wolfe2014")+
  ggtitle("Simulated time series, 3 replicates per species")

```



```{r}
ggplot(data = dat.summary,
       aes(x = name, y = count, fill = name))+
  geom_col()+
  facet_wrap(~ day)+
  xlab("")+
  scale_fill_wolfe(dat.names = dat$name, palette.name = "wolfe2014")+
  ggtitle("Simulated summary data (by day of experiment)")

```

TECHNICAL NOTE: these two functions are mostly just wrappers for `scale_fill_discrete()` and `scale_color_discrete()`, and so can be treated just like any other ggplot layer, including adding to existing objects, etc.

## What if I'm working with some new taxa but still want to use a palette for the remaining taxa?

`scale_fill_discrete()` and `scale_color_discrete()` both have an optional argument, `augment`, that allows you to provide a data frame with the new taxa and corresponding colors. As this is likely to be a somewhat iterative process (checking to see if you have all the taxa represented, adding a new taxa, checking again), `palette_augment_helper()` assists in testing your augments to see if they are sufficient or if there are redundancies between them and the palette. For the iterative process, you will want to specify the list of names in the data, provide the augment data frame (if you have it, but if you're just starting you can not specify it, and the function will list the missing colors), and provide the palette name. If you find you don't want the plot of the augmented palette with each iteration, or don't want the list of all taxa already covered in the palette, you can set `plot.palette = FALSE` and `spellcheck = FALSE`.

Here's an example, with some simple fake data that includes taxa present in the "wolfe2014" palette, and new taxa

```{r}
# set up example data. here we only care about the column with our names that
# we want to base our colors off of
dat = data.frame(spec = c("arthrobacter",
                          "yaniella",
                          "vibrio",
                          "asclepias",
                          "drosophila"),
                 vals = rnorm(5))

## first attempt at our augment:
aug = data.frame(name = c("asclepias", "vibrio"),
                 color = c("coral", "blue"))

palette_augment_helper(dat.names=dat$spec,
                       palette.name = "wolfe2014",
                       augment = aug)
## oops! Palette already has vibrio! And we're missing drosophila

aug = data.frame(name = c("asclepias", "drosophila"),
                 color = c("coral", "purple"))

palette_augment_helper(dat.names=dat$spec,
                       palette.name = "wolfe2014",
                       augment = aug)
## Looks like we're good!

ggplot(dat, aes(x = spec, y = vals, fill = spec))+
  geom_col()+
  scale_fill_wolfe(dat$spec,
                   palette.name = "wolfe2014",
                   augment = aug)


```


## adding new color palettes

It is very easy for me to add new color palettes to the package if I'm given 
-  the name for the new color palette
-  a list of the taxa names 
-  a list of the corresponding colors, in R readable format (either character strings for R colors, or in hex code (the format that starts with a # sign.. I found that [this site](https://colors.artyclick.com/) was handy for extracting colors from publications. 

If you're developing a palette from scratch for your data set, the palette_augment_helper() function can again be helpful. If you don't give it a `palette.name` argument, it just compares the augment to the data and identifies missing colors (plus plots the color scheme of the augment). I find this helpful for iteratively building a new palette for my data.
